% `phylobase`
% François Michonneau - @FrancoisInvert
% iEvoBio -  June 24<sup>th</sup>, 2014

```{r, include=FALSE}
options(markdown.HTML.header = "extra.css")
```

# Contributors

- Ben Bolker
- Marguerite Butler
- Peter Cowan
- Damien de Vienne
- Dirk Eddelbuettel
- Mark Holder
- Thibaut Jombart
- Steve Kembel
- David Orme
- Brian O'Meara
- Emmanuel Paradis
- Jim Regetz
- Derrick Zwickl
- **YOU?**

-------

## What `phylobase` is not

- a data analysis package

-------

## What `phylobase` provides

- classes + methods for object manipulation
- object = phylogenetic tree (+ data)
- robust
    - over 200 unit tests
	- S4 class
	
-------

## Why `phylobase`?

- `phylo4` as the unifying standard for PCM in `R`

- Goals
    - common data structure
	- common base functions
	- associate tree and comparative data
	- easy to manipulate

-------

## Why `phylobase`?

- For users
    - pass data from one package to another
    - keep data and tree synchronized
   
- For developers
    - focus on method development

-------

## The `phylo4` class

```{r, echo=1:2, results='hide', message=FALSE, prompt=TRUE}
library(phylobase)
data(geospiza)
data(geospiza_raw)
geosp <- phylo4(geospiza_raw$tree)
```

```{r,echo=FALSE}
geosp
```

--------

## The `phylo4` class

### Plotting

```{r, prompt=TRUE, caption='phylobubble plot'}
plot(geospiza)
```

--------

## The `phylo4` class

### Basic information about the tree

```{r, prompt=TRUE}
nTips(geospiza)
tipLabels(geospiza)
nodeLabels(geospiza)
isUltrametric(geospiza)
```

------

## The `phylo4` class

### Tree-walking

```{r, prompt=TRUE}
descendants(geospiza, 21)
ancestors(geospiza, "pauper")
```

-------

## The `phylo4d` class

```{r, include=FALSE}
geoTr <- geospiza_raw$tree
geoData <- geospiza_raw$data
```

### Put together tree and data

```{r, prompt=TRUE}
geo <- phylo4d(geoTr, geoData)
geo <- phylo4d(geoTr, geoData, missing="OK")
head(geo, n=10)
```

--------

## Other functions

- flexible subsetting functions
- options to control object validity (useful for developers)

```{r, prompt=TRUE}
phylobase.options("allow.duplicated.labels")
phylobase.options("allow.duplicated.labels" =  "ok")
phylobase.options("allow.duplicated.labels")
```

--------

## Importing data

### from NEWICK

--------

## Importing data

### from NEXUS files (using NCL)

```
#NEXUS

BEGIN TAXA;
	DIMENSIONS NTAX=18;
	TAXLABELS
		Myrmecocystuscfnavajo Myrmecocystuscreightoni Myrmecocystusdepilis Myrmecocystuskathjuli Myrmecocystuskennedyi Myrmecocystusmendax Myrmecocystusmexicanus Myrmecocystusmimicus Myrmecocystusnavajo Myrmecocystusnequazcatl Myrmecocystusplacodops Myrmecocystusromainei Myrmecocystussemirufus Myrmecocystussnellingi Myrmecocystustenuinodis Myrmecocystustestaceus Myrmecocystuswheeleri Myrmecocystusyuma 
	;
END;

BEGIN CHARACTERS;
	TITLE  Foraging;
	DIMENSIONS  NCHAR=2;
	FORMAT DATATYPE = STANDARD GAP = - MISSING = ? SYMBOLS = "  0 1 2";
	CHARSTATELABELS 
		1 time /  diurnal crepuscular nocturnal, 2 subgenus /  Endiodioctes Eremnocystus Myrmecocystus ; 
	MATRIX
	Myrmecocystuscfnavajo    22
	Myrmecocystuscreightoni  11
	Myrmecocystusdepilis     00
.......
;
END;

BEGIN TREES;
	TRANSLATE
		1 Myrmecocystuscfnavajo,
		2 Myrmecocystuscreightoni,
		3 Myrmecocystusdepilis,
		4 Myrmecocystuskathjuli,
		5 Myrmecocystuskennedyi,
		6 Myrmecocystusmendax,
		7 Myrmecocystusmexicanus,
		8 Myrmecocystusmimicus,
		9 Myrmecocystusnavajo,
		10 Myrmecocystusnequazcatl,
		11 Myrmecocystusplacodops,
		12 Myrmecocystusromainei,
		13 Myrmecocystussemirufus,
		14 Myrmecocystussnellingi,
		15 Myrmecocystustenuinodis,
		16 Myrmecocystustestaceus,
		17 Myrmecocystuswheeleri,
		18 Myrmecocystusyuma;
	TREE bestML = (((((((((13:1.724765,11:1.724765):2.926053,6:4.650818):0.689044,(4:1.08387,17:1.08387):4.255993):0.198842,((8:2.708942,3:2.708942):2.027251,((12:2.193845,10:2.193845):2.257581,18:4.451425):0.284767):0.802512):0.506099,5:6.044804):4.524387,2:10.569191):0.836689,(14:2.770378,15:2.770378):8.635503):0.89482,16:12.300701):1.699299,(7:5.724923,(1:2.869547,9:2.869547):2.855375):8.275077);
END;
```

------

## Importing data

### from NEXUS files (using NCL)

```{r, prompt=TRUE}
myrme <- readNexus(file="treeWithDiscreteData.nex")
head(myrme, n=10)
```

--------

## Importing data

### from NeXML files (using RNeXML)

```{r, eval=FALSE, prompt=TRUE}
library(RNeXML)
nxml <- nexml_read("http://purl.org/phylo/treebase/phylows/study/TB2:S15218?format=nexml")
```

```{r, include=FALSE}
library(RNeXML)
nxml <- nexml_read("phyrella.xml")
```

```{r, prompt=TRUE}
phyrella <- phylo4(nxml)
```

```{r, echo=FALSE}
head(phyrella, n=5)
```

```{r, prompt=TRUE, caption=FALSE}
plot(phyrella)
```

---------

## Future

- Support for `multiPhylo4` and `multiPhylo4d` classes
- Better/faster underlying data structure for the trees

---------

## More information

- [http://github.com/fmichonneau/phylobase](http://github.com/fmichonneau/phylobase)
- [http://cran.r-project.org/package=phylobase](http://cran.r-project.org/package=phylobase)
- Vignette for more examples
